Online {\IterComp}

### 2017-09-15 results.tex
changes to the result section; and other minor changes

We first compare static aspects of the instrumentation algorithms. The naive instrumentation instrument all basic blocks, by definition.
First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG. Afterwards, we compare these profiling
techniques with the whole program relaxation.

Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.

%\begin{figure*}[ht]
%    \centering
%    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
%    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
%    \label{fig:num-probes}
%\end{figure*}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
    \label{fig:overhead-O3}
\end{figure*}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
    \label{fig:error-O3}
\end{figure*}

For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimization.
For each benchmark, we compute the average overhead over all its 1000 input dataset.
When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.

The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
This means that the experiments show its best performance enabled by having perfect profiling information.

%\begin{figure*}[ht]
%    \centering
%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
%             Notice that the average of the ratios does not equal the ratio of the averages.}
%    \label{fig:overhead-improvement-O3}
%\end{figure*}

%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation,
%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.

Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
while the WPO relaxation is significantly less conservative.
Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.

@@

For comparison, we use five configurations.
In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}[leftmargin=3mm]
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}
The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.

@@

In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
For each benchmark, after selecting the final optimization sequence, we compute the average speedup over all the 1000 input dataset.
When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
This figure shows that the best optimization sequence selected with the Oracle-PP is very close to the performance of the best optimization sequence selected with the Oracle-RM.
This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimization search,
it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

>>>>

In this section we evaluate the performance overhead incurred by the work efficiency profiling.
We measure the wall-clock time of the benchmarks when compiled with the {\flagstype -O3} optimization.
For each benchmark, we compute the average overhead over all its 1,000 inputs.
Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work
profiling with each one of the instrumentation strategies, compared to their
non-instrumented counterparts.
The WPO relaxation was performed using profiling information for every specific
input, which provides the perfect information to perform the WPO relaxation.
This means that the experiments show its best performance enabled by having
perfect profiling information.

%\begin{figure}[t]
%    \centering
%    \includegraphics[width=0.395\textwidth]{figs/adpcm_d-cfg-instr.pdf}
%    \caption{CFG of the function that contains the hot loop of the {\flagstype adpcm\_d} benchmark.}
%    \label{fig:adpcm_d-cfg-instr}
%\end{figure}

While the optimal profiling has a maximum overhead of almost 60\%,
the work profiling with a relaxation of 5\% threshold has an overhead of less
than 16\% for all benchmarks,
with the WPO relaxation reaching an average overhead of only 5\%,
compared to an average of 12\% for the optimal profiling.

\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
    \label{fig:overhead-O3}
\end{figure*}
\begin{figure*}[t!]
    \centering
    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
    \label{fig:error-O3}
\end{figure*}

The most critical case, amongst the evaluated benchmarks, happens with the
\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
instrumentation.
With a threshold of only 5\%, the relaxation reduces
the overhead by a factor of $4.5\times$ over the optimal instrumentation,
and the WPO relaxation by a factor of $4.8\times$.
This benchmark has a single hot function
which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
Most of the improvement achieved by both strategies comes from removing only two
probes from the hot loop.
The two relaxed probes were placed in branches with a high probability of being
taken (above 55\% probability), but with a small contribution to the total amount
of work measured in DAG for this loop (both probes with an static error of
about 1.3\%).
Because these probes were placed in basic blocks with a considerable execution
frequency, their removal resulted in a significant reduction in performance
overhead.

A few of the benchmark present abnormal regressions in their execution time.
This increase in the profiling overhead is counter-intuitive because the
relaxation algorithms only reduces the number of instrumented probes,
without changing their placement.
We suggest some reasons for this abnormal regressions.
By analysing these cases, we noticed that several of the frequently executed
probes could not be removed by the relaxation.
Moreover, another important aspect to point out is that we also observed
differences among the final CFGs after applying the \texttt{-O3} optimisation.
Although relaxation does not directly alter the CFG, it is worth mentioning that
many optimisation passes make use of LLVM's built-in target-based cost model,
which means that adding or removing instructions from basic blocks may affect
decisions during transformations of the code.
For example, some of the optimizations that make use of this built-in cost model
and also alter the CFG are: loop-unrolling, function inlining, and
simplifications of the CFG.
These changes in the program's CFGs may affect other compiler optimisations and
also possibly in hardware performance, for example affecting the instruction
caching.

Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
while the WPO relaxation is significantly less conservative.
Although the error bound is not guaranteed by the WPO relaxation,
its dynamic error was always below the 5\% thresholds due to the use
of perfect profiling information.

@@

For comparison, we use five configurations.
In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}[leftmargin=3mm]
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}

The comparison between Oracle-RM and Oracle-PP is crucial for validating the
effectiveness of the work efficiency metric for guiding online {\itercomp},
while the other configurations demonstrate the viability of applying online
{\itercomp} in real-world scenarios.
For the relaxation strategies, we evaluate only those with a 5\% threshold, since
they offered better overhead reductions for little dynamic error in the measurement.

@@

In order to evaluate the quality of the final optimization sequences selected by
each configuration of the online {\itercomp}, we compare their speedup over the
{\flagstype -O3} optimization across all the 1,000 inputs of the benchmark
being optimized.
%When measuring the wall-clock time for each input, to reduce noise, we execute
%the same input until we have a statistically sound measurement, i.e. we execute
%until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups for all test benchmarks.
This figure shows that the best optimization sequence selected with the Oracle-PP
is very close to the performance of the best optimization sequence selected with
the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
improved obtained by Oracle-RM.
This result is important for demonstrating that a work-based metric has the
potential to produce good results in a real-world online scenario.

Moreover, the evaluation also indicates that the use of relaxation algorithms
does not degrade the optimization search, it might even benefit the {\itercomp}
as it tends to insert smaller interferences in the performance measurement.
%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
